---
title: "Getting & Cleaning Data -- Course Project"
output: html_document
---

<!-- For more info on RMarkdown see http://rmarkdown.rstudio.com/ -->

<!-- Enter the code required to load your data in the space below. The data will be loaded but the line of code won't show up in your write up (echo=FALSE) in order to save space-->
```{r echo=FALSE}
setwd("C:/Users/User/rpractice/GettingAndCleaningData/UCI HAR Dataset")
```
Set your working directory to UCI HAR Dataset.
<!-- In the remainder of the document, add R code chunks as needed -->

### Step 1: Merge the training and test sets to create 1 data set:

####Preliminary steps:

The packages we will need to complete this project:

```{r}
library(data.table)
```  

```{r}
library(dplyr)
```


####Working with the data:

From looking at the files in the UCI HAR Dataset directory, we can divide the pertinent files into three main groups:  

1.**Subject**: values come from subject train and test files.  

2.**Activity**: values come from X train and test files.  

3.**Features**: values come from Y train and test files.  


Names of variable Features come from features.txt file, and levels of variable Activity come from activity_labels.txt files.  

To complete the task outlined in this step, we will do the following:  

####Read the datasets into R:

**Subject**:

```{r}
subjectTrain<-read.table("train/subject_train.txt")
```
```{r}
subjectTest<-read.table("test/subject_test.txt")
```

**Activity**:

```{r}
activityTrain<-read.table("train/y_train.txt")
```
```{r}
activityTest<-read.table("test/y_test.txt")
```

**Features**:

```{r}
featuresTrain<-read.table("train/X_train.txt")
```
```{r}
featuresTest<-read.table("test/X_test.txt")
```

####Read the labels/names into R:

```{r}
activity_labels<-read.table("activity_labels.txt", header=FALSE)
```
```{r}
feat_names<-read.table("features.txt")
```

####Merge the subject (train & test), Activity (train & test) and Features (train & test) data using rbind():

```{r}
subject<-rbind(subjectTrain, subjectTest)
```
```{r}
activity<-rbind(activityTrain,activityTest)
```
```{r}
features<-rbind(featuresTrain,featuresTest)
```

####Set names for the variables in the resulting datasets:

```{r}
names(subject)<-"subject"
```  

```{r}
names(activity)<-"activity"
```  

```{r}
names(features)<-as.character(feat_names[,2])
```


####Finally, merge the resulting three datasets into one using cbind():  

```{r}
data<-cbind(subject,activity, features)
```

####Clean the unnecessary data from the workspace:

```{r}
rm(subjectTrain,subjectTest,activityTrain,activityTest,featuresTrain,featuresTest, subject,activity,features)
```



### Step 2: Extract only the measurements on the mean & SD for each measurement:

Column Indices for columns with means and standard deviations:

```{r}
Means<-grep("mean()", colnames(data))
```  

```{r}
SDs<-grep("std()", colnames(data))
```  

```{r}
MeansAndSDs<-c(Means,SDs)
```

Extract the above-indiced columns (while keeping the variables "subject" and "activity" in columns 1 and 2, respectively):

```{r}
newdata<-data[, c(1,2,MeansAndSDs)]
```

Let's take a look at str(newdata):

```{r}
str(newdata)
```

We can remove the columns with the meanFreq measurements:

```{r}
NewData<-newdata[, !grepl("Freq", colnames(newdata))]
```

We currently have a dataset with only the measurements on the mean and SD for each measurement.

### Step 3: Use descritive activity names to name the activities in the dataset:

To do that, we will be using the file activity\_labels.txt (which we have preveiously read into activity_labels dataframe).  

Change variable activity from numeric to character so that we could use the activity names:

```{r}
NewData$activity<-as.character(NewData$activity)
```
  
  
"Attach" activity names from the file activity labels.txt:

```{r}
for (i in 1:6){NewData$activity[NewData$activity==i]<-as.character(activity_labels[i,2])}
```

Factorise variable activity:

```{r}
NewData$activity<-as.factor(NewData$activity)
```

Check:

```{r}
 class(NewData$activity)
```  

```{r}
head(NewData$activity, 10)
```


### Step 4: Appropriately label the dataset with descriptive variable names:

Let's take a look at our variable names:

```{r}
names(NewData)
```
  
  
Having examined our variable names, we can make the following changes:

"t" in the beginning of the variable name, to "time":

```{r}
names(NewData)<-gsub("^t", "time", names(NewData))
```

"f" in the beginning of the variable name, to "frequency":  

```{r}
names(NewData)<-gsub("^f", "frequency", names(NewData))
```

"Acc" to "Accelerometer":  

```{r}
names(NewData)<-gsub("Acc", "Accelerometer", names(NewData))
```

"Gyro" to "Gyroscope":  

```{r}
names(NewData)<-gsub("Gyro", "Gyroscope", names(NewData))
```

"Mag" to "Magnitude":

```{r}
names(NewData)<-gsub("Mag", "Magnitude", names(NewData))
```

and finally, "BodyBody" to "Body":  

```{r}
names(NewData)<-gsub("BodyBody", "Body", names(NewData))
```

Let's take another look at our variable names after the change:

```{r}
names(NewData)
```
  
  
They are now significantly more descriptive than they were when they were abbreviated.


### Step 5: From the dataset in step 4, create a 2nd, tidy dataset with the average of each variable for each activity and each subject:

Creating a tidy dataset with the average of each variable for each activity and each subject, using the aggregate() function:

```{r}
TidyData<-aggregate(.~subject+activity, NewData, mean)
```

Creating a txt file containing our tidy data set, using the write.table() function from the data.table package:

```{r}
write.table(TidyData, file="C:/Users/User/rpractice/GettingAndCleaningData/tidy.txt", row.names=FALSE)
```

The resulting dataset satisfies the conditions for a tidy dataset:

  * Each variable we measured is in one column (see codebook for a complete list of all 68 variables).  
  * Each different observation forms a row (we have 30 subjects and 6 levels of activity (laying, sitting, standing, walking, walking downstairs & walking upstairs): a total of 180 rows of observations). 
  * Each type of observational unit forms a table.
  
